Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan

### 2019-11-19 analysis/paper/paper.Rmd

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "glass bead", "medium bead", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artefact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)
```

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares were excavated _in situ_ from household area of Kiwulan, indicated by numerous post-holes. Thirty burials contained 3173 ornaments, and 27 ornaments were recovered from middens. In the burials ornaments were in the form of bead strings that consisted of hundreds to thousands of pieces. The ornaments from the burial contexts are not analysed here since few burials can be confidently assigned to a time period. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organisation. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

@@

# Results of hypothesis testing

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequency for the major class of ornaments between different time periods at Kiwulan. The result of chi-square test for the number of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$) shows that there is a non-random difference between the three time periods. It indicates that most ornament types were present before European contact. Ornament frequencies reached a peak during European period and then dropped during the Chinese period, especially for golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period.

@@

The distribution of frequency for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. According to the frequency, agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the Europeans. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in a cluster that indicates the original use as strings or necklaces. The common shape for copper bells is the large shape with a wide variety of human faces as a motif, while the common shape for glass beads is a bead with length less than 1 cm that are often identified as Indo-Pacific beads, the most common type that had been widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all ornaments from the research area for each time period. Before the European arrival, a greater amount of ornaments were found at the northern and middle parts of the research area. During the European period, ornaments were more widespread, with some clusters on the northern part. During the Chinese period the distribution is more even again. Figure \@ref(fig:plot-spatial-pattern-each) presents the distribution for the major ornament classes individually, some clusters across the area can be observed during the European period, such as golden beads and agate beads. However, there seems to be no clear consistent pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, the cluster of golden beads was found at the northern part, while the cluster of agate beads was found in the middle part. In contrast, there are multiple clusters of metal rings that are distributed separately across the research area. Copper bells were usually found individually and seem randomly distributed across the area. In the Chinese period, both the amount and density of different classes of ornaments decreased.

@@

I used point pattern analysis to examine whether the distribution of ornaments was uneven in non-random ways that could reflect changes in social inequality. Point pattern analysis is a way to assess whether the distribution of sites or artefacts represent some “hotspots” produced by non-random processes, implying the influence of social processes [@Bevan2016; @Ducke2015]. Hotspots in the pattern of ornaments could indicate changes in social organisation due to culture interaction. To explore ornament distributions I used the intensity approach that focuses on the average density of points across space. During excavation, ornament locations were only recorded to the square (4 x 4 m). To prepare the location data for point pattern analysis, I assigned each ornament to a random coordinate pair located in square it was recovered from. The next step was to subset the ornaments into three groups for three time periods. Finally I computed the kernel densities for each time period for comparison.

Kernel density estimations (KDE) estimate the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here I use KDE to visualize core areas of ornaments and surrounding neighborhoods [@Bonnier2019; @Cortegoso2016]. Density values of artefacts per square meter were calculated for each cell.

The results show that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period according to the kernel density maps (Figure \@ref(fig:plot-all-densities)). It seems that there are three consistent sub-regions with a core area that shifts over time. The distribution might indicate different household-based social groups who possessed more ornaments. The multiple groups during European period might reflect more unequal consumption of ornaments across the site, relative to other periods, to indicate social processes caused by human behavior. However, it could also just a random pattern resulting from bigger sample size. In addition, the generation of core areas might be biased due to small sample sizes, for example, a few ornaments found at one single square during the Chinese period could create an obvious hot spot. To further evaluate these patterns, we test the hypothesis that the distribution of ornaments is not randomly distributed to determine if social processes might explain ornament distributions.

@@

To test this hypothesis about the randomness of spatial locations, I used a Monte Carlo method to simulate average nearest-neighbor distances (ANN). Our null hypothesis is that the distribution of ornaments is consistent with a completely random process. We simulated the locations of ornaments across the space 1000 times for each time period. The observed data was compared to the random patterns generated by simulated processes to determine if our data are randomly distributed or not. Figure \@ref(fig:hypo-test-distribution) shows the distributions of the ANN distances calculated on 1000 simulations of ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European period, which means the ornaments have non-random clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese period, with about one third of the simulated values are greater than our observed ANN value. This testing reveals that the clustered distributions of ornaments during the pre-European and European periods might reflect the presence of different social groups. Moreover, the clustering of ornaments during the European period is highly non-random, might potentially indicating that some of the social changes leading to different degrees of access foreign ornaments or a concentration of power to control the distribution of ornaments at Kiwulan.

@@

The results show there was a greater diversity of ornaments types and materials at Kiwulan during the European presence. This might result from the large scale exchange network that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the mortuary practices of Han Chinese. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007].

Spatial patterns of ornaments shows that their distribution was clustered during the pre-European and European period. Hypothesis testing for spatial patterns indicates that these clusters are non-random, and are most highly concentrated during the European period. Because ethnographic evidence indications that trade ornaments were prestige goods in prehistoric Northeastern Taiwan, this spatial clustering may indicate that a degree of social inequality was present before European contact and then it was reinforced and amplified during the European period. In addition, there was a burial dated to 17th century unearthed 60 golden beads above the average of 2-3 pieces in the pre-European period, further indicating increasing social inequality [@Chen2007; @Cheng2008].

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. For example, the corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labor tasks [@Feinman2000; @Siegel1999].

We may consider that Yilan social organisation went from corporate mode before the European arrival, then transitioned into network mode during the European presence, then back to corporate mode in the Chinese period. The small number of ornaments, and less concentrated distribution during the pre-European period is consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by the Europeans resulted in the appearance of a network mode due to competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, it implies that local leaders might have flexibility to manipulate European colonial image, expand personal power, and monopolize the high-value trade goods [@Kang2012]. During the Chinese period, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This reflects the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals accumulate wealth.

@@

Examination of the indirect influence of colonialism can reveal how peripheral areas were influenced by the colonial activities or involved in the colonial economy centred in the major European colonies [@Trabert2017]. Kiwulan in Northeastern Taiwan is an important case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan can provide valuable insights into the discussion of indirect colonial influence on the local societies living beyond the borders of direct European colonial occupation. The frequency and spatial distribution of body ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects international exchanges in a colonial context. Ornaments were viewed as prestige goods in local indigenous culture before European contact. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may be intensified that further signals wealth and privileged trading connections with European colonists for the inhabitants of Kiwulan. These symbolic values may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality.

This study demonstrates that foreign ornaments can be a proxy to detect indirect colonial influence on local indigenous populations. Ornaments can give insights into the amplification of social inequality stimulated by European colonisation. It also shows the agency of indigenous people to incorporate ornaments into their social system and use them in their daily lives to display or intensify status differences. Future work could extend this approach to studies of other trade goods such as ceramics. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and future tests of this should include analysis of pottery production and standardisation, and mortuary practices.

>>>>

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "glass bead", "medium bead", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artefact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)

#
golden_bead_n <- 
  ornaments_count %>% 
  filter(Categories == "Golden bead") %>% 
  pull(n)

glass_bead_n <- 
  ornaments_count %>% 
  filter(Categories == "Glass bead") %>% 
  pull(n)

agate_bead_n <- 
  ornaments_count %>% 
  filter(Categories == "Agate bead") %>% 
  pull(n)

metal_ring_n <- 
  ornaments_count %>% 
  filter(Categories == "Metal ring") %>% 
  pull(n)

metal_bell_n <- 
  ornaments_count %>% 
  filter(Categories == "Bell") %>% 
  pull(n)
```

@@

This study focuses on `r sam_post_hole` ornaments from 40 sampling squares that were excavated _in situ_ from the household area of Kiwulan, indicated by numerous post-holes (Figure \@ref(fig:KWL-sam-area)). Thirty burials contained 3,173 ornaments, and 27 ornaments were recovered from middens. In the burials ornaments were in the form of bead strings that consisted of hundreds to thousands of pieces per burial. The ornaments from the burial contexts are not analysed here since few burials can be confidently assigned to a time period. The household area represents domestic spaces suitable for testing for changes in ornament distribution relating to social organisation. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament are golden beads (n = `r golden_bead_n`), followed by glass beads (n =  `r glass_bead_n`), agate beads (n = `r agate_bead_n`), metal rings (n = `r metal_ring_n`), and metal bells (n = `r metal_bell_n`, see also Figure \@ref(fig:ornaments-image)). The variety of subtypes and their individual frequencies are summarized in Table \@ref(tab:tidy-data-for-table).

@@

# Results

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequencies of the major classes of ornaments for different the time periods at Kiwulan. The difference in frequencies between the three time periods reflect significant differences in the use of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$). Most ornament types were present before European contact. Ornament frequencies reached a peak during the European period and then dropped during the Chinese period, especially golden beads. This trend can be also seen on other ornaments including agate beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period.

@@

The distribution of frequencies for subtypes in each major class are presented in Figure \@ref(fig:plot-altogether). Spearman's correlation test shows that there is no significant relationship between diversity of subtypes and sample size (S = `r spearman_result_stat`, rho = `r spearman_result_rho`, p = $`r spearman_result_p_value`$). This indicates that the increases in diversity can be explained by the effects of culture interaction instead of effects of sample size. Agate beads and metal rings have greater quantity and variety of shapes compared to copper bell and glass beads during the European period. The greater varieties for agate beads and metal rings might indicate multiple origins due to participation in global trade networks stimulated by the European presence. In contrast, copper bells and glass beads have less variety, but glass beads have a larger number because they were usually found in clusters that indicate their original use as strings or necklaces. Copper bells are typically >2 cm long with a wide variety of human faces as a motif, while glass beads are usually <0.5 cm long Indo-Pacific beads, widespread in Southeast Asian sites since 300 BC [@Francis2002].

@@

Figure \@ref(fig:plot-all-spatial-pattern) presents the spatial distribution of all ornaments from the research area for each time period. Before the European arrival, a greater amount of ornaments were found at the northern and middle parts of the research area. During the European period, ornaments were more widespread, with some clusters on the northern part. During the Chinese period the distribution is more even again. Figure \@ref(fig:plot-spatial-pattern-each) presents the distribution for the major ornament classes individually, some clusters across the area can be observed during the European period, such as golden beads and agate beads. However, there seems to be no clear consistent pattern across those different ornaments. Each class shows its own pattern where the squares with higher number of ornaments distributed separately and independently. For example, a cluster of golden beads was found at the northern part, while a cluster of agate beads was found in the middle part. In contrast, there are multiple clusters of metal rings that are distributed separately across the research area. Copper bells were usually found individually and appear randomly distributed across the area. In the Chinese period, both the amount and density of different classes of ornaments decreased.

@@

Point pattern analysis can assess whether the distribution of artefacts represents hotspots produced by non-random processes [@Bevan2016; @Ducke2015], such as concentrations of ornaments resulting from social inequality stimulated by a colonial presence. To prepare the ornament location data for point pattern analysis, I assigned each ornament to a random coordinate pair located in 4 x 4 m square it was recovered from (ornaments lack individual point provenance). The next step was to divide the ornaments into three time periods. Finally I computed the kernel densities for each time period for comparison. Kernel density estimations (KDE) compute the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here I use KDE to visualize core areas of ornaments and surrounding neighbourhoods [@Bonnier2019; @Cortegoso2016]. Density values of artefacts per square meter were calculated for each cell.

Figure \@ref(fig:plot-all-densities) shows that there is one major core area during the pre-European period, multiple core areas during European period, and a single core during the Chinese period. There are three consistent sub-regions with a core area that shifts over time. The distribution might indicate increase and decrease in the number of different household-based social groups who possessed more ornaments. The multiple groups during the European period might reflect more unequal consumption of ornaments across the site, relative to other periods, or random patterns resulting from bigger sample size. In addition, the generation of core areas might be biased due to small sample sizes, for example, a few ornaments found at one single square during the Chinese period could create an obvious hotspot.

@@

To test for randomness in spatial locations, I used a Monte Carlo method to simulate average nearest-neighbour distances (ANN). Figure \@ref(fig:hypo-test-distribution) shows the observed ANN distances with the distributions of the ANN distances calculated on 1000 simulations of random ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European period, which means the ornaments have non-randomly clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese period, with about one third of the simulated values are greater than our observed ANN value. This testing reveals that the clustered distributions of ornaments during the pre-European and European periods might reflect the presence of different social groups. Moreover, the clustering of ornaments during the European period is highly non-random, potentially indicating different degrees of access to foreign ornaments or a concentration of power to control the distribution of ornaments at Kiwulan.

@@

An indirect colonial presence is indicated at Kiwulan by a greater diversity of ornaments types and materials during the European period. This likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. However, the frequency of overall ornaments and each subtype declines significantly during the Chinese period. This may be due to a smaller scale of trading network, the overall decline of the indigenous population in Yilan, and adopting the practices of Han Chinese who did not use these kinds of beads in their dress. Since the end of the 18th century, many indigenous people moved southwards to Hualien due to population pressure caused by Han Chinese immigrants [@Chen2007].

Spatial patterns of ornaments shows that their distribution was clustered during the pre-European and European period. Testing for spatial patterns indicates that these clusters are non-random, and are most highly concentrated during the European period. Because ethnographic evidence indicates that trade ornaments were prestige goods in prehistoric Northeastern Taiwan, this spatial clustering may indicate that a degree of social inequality was present before European contact and then it was reinforced and amplified during the European period. A further indicator of increasing social inequality is a burial dated to the 17th century that included 60 golden beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008].

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. For example, the corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999].

Yilan social organisation went from corporate mode before the European arrival, then transitioned to network mode during the European presence, then back to corporate mode in the Chinese period. The small number of ornaments, and less concentrated distribution during the pre-European period is consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode due to competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeast Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012]. During the Chinese period, the decreasing frequency and diversity of ornaments show a decline in the production, use and discard of ornaments at Kiwulan. This reflects the smaller scale of exchange networks, with limited sources of ornaments, and limited opportunities for individuals to accumulate wealth.

@@

Examination of the indirect influence of colonialism can reveal how peripheral areas were influenced by colonial activities or involved in the colonial economy centred in the major European colonies [@Trabert2017]. Kiwulan in northeast Taiwan is an exceptional case study as an East Asian location that was relatively isolated and peripheral, and yet connected by regional and global trade networks. Kiwulan provides valuable insights into the discussion of indirect colonial influence on local societies living beyond the borders of direct European colonial occupation. The frequency and spatial distribution of body ornaments at Kiwulan present three distinct patterns during different dominant culture interaction periods. The greater amount and diversity of ornament types during the European period reflects international exchanges in a colonial context. Ornaments were viewed as prestige goods in the local indigenous culture before European contact. After the arrival of the Europeans, the exotic and powerful image carried by those ornaments may have intensified, further signalling wealth and privileged trading connections among the inhabitants of Kiwulan. These symbolic values may have stimulated more competition between aggrandizing individuals for prestige and wealth accumulation at Kiwulan, which might have resulted in an increase in social inequality.

We are still far from understanding the full variety of colonial impacts on peripheral indigenous communities. The Kiwulan ornaments suggest that foreign ornaments can be a proxy to detect indirect colonial influence on local indigenous populations. Ornaments give insights into the amplification of social inequality stimulated by European colonisation. It also shows the agency of indigenous people to incorporate ornaments into their social system and use them in their daily lives to display or intensify status differences. Future work could extend this approach to studies of other trade goods such as ceramics. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and future tests of this should include analysis of pottery production and standardisation, and mortuary practices.
